\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}

\title{Preregistering A Project Evaluating Social Media Risk Communication for Vaccine Safety}
\author{David Manheim, Anat Gesser-Adelsburg}
\date{May 2019}

\begin{document}

\maketitle

\section{Introduction}
Risk Communication is a difficult but critical task in public health. The task of risk communication on social media is particularly challenging because of the rapidly changing nature of the platform, and the challenges of active engagement with the public. 

Twitter is being used extensively in various positive capacities in public health\cite{Denecke2013,Paul2011, MORE}, but also serves as a location for misinformed or ill-intentioned individuals to spread misinformation\cite{Aymanns2017, Broniatowski2018, MORE}. This is particularly true in the realm of disinformation and attempts to correct it regarding vaccine safety, and Twitter has therefore become one of several battlegrounds, and has been a source of disinformation and confusion\cite{Broniatowski2018, MORE}, and risk communication for vaccines is made both more important, and more challenging.

While we hope that public health organizations are winning this battle, and know that they are at least are engaged in critical efforts to that end, success will require evaluation and adjustment of strategies. Despite recommendations about how best to use Twitter for health communication, and discussion about how these best practices are relevant to vaccine communication, not enough has been done to evaluate the use of Twitter by experts in this particular area. This type of evaluation is critical for assisting organizations in refining their strategies and further improving public health response. To support this, this research project hope to begin a process of better evaluating current efforts to inform the public and combat misinformation.

\subsection{Status of the preregistration document}
This document is both a public preregistration, and an initial draft which will be rewritten into the intended paper. The living version of the document will continue to be rewritten, but will remain available in the code repository, including history and revisions.

\section{Contribution and Extant Literature}

This paper is the first, to our knowledge, that attempts to assess how well experts involved in countering anti-vaccine propaganda interact with the general public on twitter. This evaluation also includes a new and potentially valuable, albeit data-intensive, approach for understanding conversations in historical twitter data. As will be discussed in research goals, this should also provide a valuable input for helping public health officials and experts better understand how past public conversations have been successful, and how they have fallen short.

\subsection{Evaluating Best Practices}

General best practices for risk communication have been promulgated by the World Health Organization, including a course on risk communication essentials \footnote{\url{https://openwho.org/courses/risk-communication/items/650J3rv4o76fglMyYghSMY}}, and a guide specifically for vaccine safety has been produced by the Council for International Organizations of Medical Sciences (CIOMS) Working Group on Vaccine Safety\footnote{CIOMS Guide to vaccine safety communication. Report by topic group 3 of the CIOMS Working Group on Vaccine Safety. Geneva, Switzerland: Council for International Organizations of Medical Sciences (CIOMS), 2018.} 

This paper addresses the gap in the literature for expert discussion of vaccine denial, but there has been some work done on evaluation in related areas. Messner et al. 2013 looked at the overall twitter usage of 20 nonprofit health organizations over a brief period in 2012, and looked at 948 tweets for use of hashtags, interaction with other twitter users, and other indications. Among other points, this paper is an early example that highlighted the importance of hashtags and user interaction in health risk communication\footnote{Look int this: "In reality, however, Muralidharan, Rasmussen, Patterson and Shin
(2011), in their study of nonprofit organizations’ engagement on social
media, found that two-way communication on Facebook and Twitter
is rare. Similar to news organizations, they merely use social media
channels to disseminate information through links. Lovejoy et al.’s
(2012) study supported these statements. "}. 

Radzikowski et al looked at twitter discussion of the MMR vaccine, focusing on geospatial characteristics, and provided descriptive characteristics that related these discussions to the ongoing narratives in different regions\cite{Radzikowski2016}. In addition to the focus on vaccines and the more recent and larger sample being considered here, the planned study takes the additional step of considering experts, rather than just health organizations or the general public, and performs evaluation of interaction rather than just description of participation in discussions.

Eckert et al. performed a literature review and meta-analysis to understand best practices for risk communication around emergencies, and to identify best platforms and practices to promote health protection and fight disinformation. Conclusions included a consensus that social media should be used to communicate, but many studies noted that the communication needed to be contextualized properly. Less directly relevant, there was consensus that  social media should be incorporated into daily operations of health education organizations. The reviewed studies also considered misinformation, and noted that in the context of disaster response, self-regulation by users (i.e. non-experts) was often effective\footnote{This stands in contrast to vaccination debates, where wisdom of the crowds seems to have failed to quash misinformation. This is likely due to non-expert's ability to personally verify misinformation during crises, compared to the need for expert evaluation of scientific claims in vaccine debates.}. Despite this, the studies recommended experts and partners assign dedicated social media officers who build trust and credibility, and are able to rapidly respond. The studies also found that hashtags, especially organic ones, were useful for communication, especially for "prompt myth-busting."\footnote{Again, the difference between crisis response, where hashtags are used extensively by people looking for information in real time, differs from how it is used for endemic and ongoing issues like vaccine denial. On the other hand, there are theoretical results showing that this type of fake news should be identifiable by simple agents (\textit{See citation, requires some assumptions/constraints I need to look into}\cite{Aymanns2017}.} Importantly, Eckert et al. noted that "studies on social media use during disasters were published demanding a new evaluation of emerging evidence on best practices for governmental agencies, implementing partners and the public." \cite{Eckert2018}

Elbanna et al. more recently looked at operational implications of social media for emergency management, and proposed a research agenda on the basis of workshops with individuals and organizations. This paper identified six challenges, "which include the dissemination of information, disaster tourism, communities’ expectations, flash volunteering, PPRR, crisis communication and community relations." \cite{Elbanna2019} Some of these issues seem likely to emerge in our investigation as well, and the relationship between the identified challenges and what we expect to find will be explored.

\subsection{Methodological Novelty}
As discussed in more detail in the methodology, the current study attempts to find all conversations that involved the experts, rather than relying on the poorly-understood Twitter 1\% sample, and previous studies have done. This is accomplished by retrieving the full (available) record of all participant conversations, and tracing the conversations from that expanded corpus. This will also potentially allow further analysis of the other participants in the conversation, to identify whether they were other experts, vaccine doubters, or otherwise.

In addition, the analysis of the tweets is intended to explore more of the conversation about vaccines, and capture context rather than just the behavior of experts and organizations.

\section{Research Goals}

Unfortunately absent from the literature were attempts to evaluate the identified best practices, and this is the key area that the current work hopes to begin to address. The significant gap in the literature on evaluating social media engagement in public health means that there is little feedback about how well organizations are faring in implementing the suggested best practices. Vaccine risk communication is both itself an important topic, and a useful case-study for how to understand and evaluate other risk communication efforts.

The primary goal of the study is to identify how organizations and individuals involved in vaccine safety risk communication are acting in practice, and to compare this to the promulgated best practices. The case study will analyze twitter accounts involved in relevant communications, collect data on their interactions, and analyzing the data to understand whether and how well the existing recommendations and best practices are followed. 

In order to ensure the analysis is valid, the research procedure is defined in this project plan, before the data is collected. This will significantly limit researcher degrees of freedom, and hopefully allow peer review of the procedure without reference to the results. \textit{TENTATIVE: To this end, open peer review of the protocol is being sought.}

\section{Analysis of Communication and Engagement Recommendations}

There are a variety of recommendation types, some of which will be more amenable to analysis than others. In addition, much of the literature on communication and response is general, rather than being specific to Twitter, so that many recommendations are inapplicable. After reviewing the literature of recommendations, the specific analytic techniques and tools, and specific metrics used for evaluation, will be explained.

\subsection{Communication Recommendations Literature Review}
The World Health Organization (WHO) provides an online course via their OpenWHO platform entitled "Risk communication essentials," available online (\href{https://openwho.org/courses/risk-communication}{LINK})\cite{OpenWHORiskComm2017}. Some of the points raised in this training are possible to evaluate, while others are not. 
 
For example, having a "Single Overarching Communication Outcome" is both hard for an external party to evaluate, and requires inferences about specific organizations' decisions and goals. Other recommended tactics, such as acknowledging the question and transitioning back to the key message, are evaluate-able but may be less useful or relevant in a length-restricted tweet. For example, recommendations such as ensuring messages start with why, then discuss how, and finally address what, are less relevant when only 280 characters are available.

Some points are easier to evaluate. The WHO training recommends that messages should "avoid jargon and technical language," that they "acknowledge uncertainty[,] mistakes [and] what you don't know," and they are respectful of concerns." Even easier to evaluate are specific rules, that messages should "avoid negative terms (not, never, no...)[, and] absolute terms (always, never, absolutely certain...)"

The WHO itself engages in outreach on social media, as they discuss on their web site\cite{WHOWebSocialMedia}, and they describe their strategy. "By participating in social media conversations and disseminating credible information through social media, WHO can drive traffic to the WHO website where more detailed and trusted content can be found." Referencing trusted sources and linking to that content is both a viable strategy on twitter, and one that can be evaluated.
\subsection{Understanding the Corpus}

Based on the WHO Risk Communications essentials course, we consider four categories of people; Champions, Silent Boosters, Avoiders, and Blockers. In the training, the WHO provides a "Quadrant Diagram" \cite{Rao2009b} where the two specta represented by the axes are Agree/Disagree and Energetic/Passive. In our case, this is Pro-vaccine, Anti-vaccine, and a spectrum of how engaged the people are on twitter. As Rao notes, this is useful for representing areas where there is ambiguity or fuzziness in (at least one of) the categories, but where despite the inability to specify an exact taxonomy, the four quadrants are different paradigms, with clear examples. The four paradigmatic examples in our analysis respectively correspond to: experts and organizations which actively promote vaccines, the non-vocal vaccination-using public, those who are unsure whether to vaccinate or quietly choose not to vaccinate, and vocal doubters and anti-vaccine activists. The primary accounts we expect to find in our corpus are the first and last group, but it is critical to notice that the most important audience is likely to be the (unobserved) third group which sees the twitter conversations\footnote{Some of these accounts might be able to be identified via likes or retweets, or at their most vocal, questions and ambivalent statements in these conversations, but these are not the focus of the corpus or the evaluations.}


\section*{To Review / Work TODO:}

The WHO recently released a best-practice guidance document on how to respond to vocal vaccine deniers in public. \url{http://www.euro.who.int/__data/assets/pdf_file/0005/315761/Best-practice-guidance-respond-vocal-vaccine-deniers-public.pdf}

{\textit WHO also recommends using strategies from "Communication for Behavioral Impact (COMBI)" in the course - I still need to find and review this.}

 { \textit WHO also provides various guidelines on their communication\cite{WHOWebHealthComm},  \url{https://www.who.int/communicating-for-health/principles/en/} }

The below table \textbf{(TODO: currently external spreadsheet, to format and include as table)} lists recommendations found in each source, and whether they are applicable or not, and if applicable, a rough estimate of how evaluate-able they are on a scale from 1, effectively impossible, to 3, possible but difficult, potentially requiring qualitative analysis of each tweet, to 5, which can be evaluated automatically with high precision.

\textit{In a 2011 article, they say \url{https://www.who.int/bulletin/volumes/89/11/11-031111/en/};
Social media and the web give people a forum to express their views, but these are debates that may never reach a conclusion. For public health specialists this can be difficult ground. “Once you start interacting with social media, you have to be ready to carry the work forward”, says Elbes of WHO. “You can't just start and stop.”}

\section{Data Sources and Data Collection Procedures}

The data for the current study will primarily consist of a corpus of tweets retrieved from Twitter. The process for defining and retreiving this corpus is somewhat complex, and decisions and procedures for data collection are critical to the validity and relevance of a study. For this reason, the procedures are outlined, and the code for retrieval and analysis will be made available along with the analysis. To the extent possible, the dataset will be made available as well.

To outline the procedure, first, a list of organizations that are involved in vaccine health education will be compiled, along with relevant twitter accounts.  Tweets from the identified accounts that mention terms used in discussion of vaccine safety and concerns about vaccine safety will be collected, along with the conversations, as described below. Depending on the cost and logistical difficulty, the initial list of accounts may be extended to both additional health education accounts, and to anti-vaccine propaganda accounts, using the procedures described below.

\subsection{Why Twitter?}
Twitter was chosen as the platform to analyze for several reasons. First, it is a widely used medium that has been noted as a focus for public discussion of vaccines. Second, it allows discussion between experts and the public, which is critical for understanding how experts engage with and respond to the public. Lastly, the data itself and the conversations are public, so that they can be analyzed.

\subsection{Data and Barriers to Access}
Twitter unfortunately restricts access to its data for academic purposes. \cite{Alaimo2018} Full access to the described dataset will be requested, and if it is possible within a reasonable budget, this will be used. Otherwise, Twitter API provides a way to query twitter and retrieve a sample of data, which can then be used to find relevant users and accounts as well as replies and discussions. 

Note that the described data collection is permitted under Twitter's policies. Tracking these groups and individuals should be allowed given that neither self-identified anti-vaccine groups nor health experts should be considered "sensitive groups and organizations." As noted below, any links between twitter and off-twitter data will be solely on the basis of claimed affiliations, and will be used only to identify whether someone is a health expert at a university, a national or international organization, or a government agency. The data will also not be used to "derive or infer potentially sensitive characteristics".

\subsubsection{Data Gathering and Ethics}
All data collection was done using the public Twitter API, and all data is accessible to the public via twitter's web site, as well as to developers via the API. All tweets that were accessed and analyzed are available publicly, and other than the World Health Organization itself, no users are explicitly identified in the discussion. The dataset used has no identification other than the data which was voluntarily published publicaly by twitter users. Other than users explicitly noting their affiliations and organizations that self-identified in the dataset, these were not further tied to non-twitter identities.

\subsection{Initial Account Lists}

The World Health Organization has a "Vaccine Safety Network", with twitter account @WHO\_VSN. This account follows a set of different account that appear to be members of this network, and these twitter accounts will form the primary list of vaccine health educators.  This includes both organizational accounts, individuals who represent organizations, and the accounts of other recognized experts who are involved in risk communication efforts and providing information about the importance of vaccination and vaccine safety.

The set of accounts that these accounts follow will be extracted to potentially supplement this list\footnote{A small selection of these accounts have been briefly inspected to ensure the reasonableness of the procedure.}. This is likely to be too extensive a list, so accounts that are "verified," are followed by multiple of these primary accounts, or are followed by a large number of verified accounts will be prioritized. Not all of these additional accounts will be relevant to vaccine risk communication, so these will be filtered. Accounts will be included if their account names or description contains terms relevant to vaccines or public health. The terms indicating account relevance are: Health, Doctor(s), Medicine, Medical, Vaccine(s), Infectious Disease(s), Immunize, and Immunization.

Depending on the size of the corpus of tweets obtained, the primary corpus may also be used to identify additional key accounts, either those that are frequently involved in the conversations, those that frequently retweet the conversations, or those that belong to celebrities or are important due to having very large numbers of followers.

In addition to these accounts, accounts involved in anti-vaccine propaganda may be examined in order to both better understand interactions, and to consider how those accounts engage with followers and other twitter users, and how they promote misinformation. This would be done in the same way the primary health education corpus is used to identify additional key accounts. Specifically, this will identify the most important twitter users that reply to discussions about vaccines and include anti-vaccine terms, as discussed below.

\subsection{Tweet and Discussion Corpus Extraction}

For each health education account, the list of tweets sent between 1/1/2018 and 1/1/2019 will be extracted. Any tweet which mentions any of the vaccine-relevant keywords or hashtags listed below will be considered a relevant tweet for the corpus. The discussions involving these tweets will then be extracted. (For example, if a tweet from an expert replies to another tweet stating that, "MMR vaccines do not contain mercury"\footnote{They do not and never have done so\url{https://www.cdc.gov/vaccinesafety/concerns/thimerosal/index.html}.}, we would retrieve not only replies to the expert's tweet, and further tweets in the thread, but also the tweet to which it replied, along with replies and tweets in that conversation.) Note that this is intended to be a very inclusive list, and is likely to include many false-positives. This ensures fewer tweets are accidentally excluded, which is critical for ensuring relevant conversations are captured.

The health education tweet term list is; Vaccine(s), Immunocompromised, Preventable disease(s), Rumors, Debunk(ed/ing), Expert(s). In addition, the terms MMR and DTaP, refer to specific, often controversial vaccines, and the terms Measles, Mumps, Rubella, Diptheria, Tetanus, Pertussis, and Whooping Cough refer to the diseases prevented.

In these discussions, we expect to find some amount of debate and anti-vaccine propaganda. The tweet term list for identifying misinformation and vaccine denial includes; Autism, Poison(s), Shills, Pharma, Mandatory, VAERS\footnote{The rare and typically minor side effects of vaccines that are collected by the Vaccine Adverse Event Reporting System (VAERS) in the United States is often cited by misinformed anti-vaccine advocates as proof that vaccines are harmful.}, Aluminum, Mercury, and Thimerosal.

The accounts involved in questioning safety and promoting anti-vaccine misinformation will be identified in part based on this term list, which may be modified based on review of the data, using the proposed analytical model discussed below.

\subsection{Iterative Retrieval Method}
Because twitter does not provide information on replies to a given user, the methodology attempts to capture the conversation "upstream" of the tweet, i.e. the tweet to which the health expert is replying. All tweets by the users replied to are retrieved, and any tweets involved in any conversations found will be extracted. For example\footnote{Neither of the twitter users used for this example currently exist.}, if expert "VaccineExpert" replies to a tweet by twitter user "AntiVaxxerMom," all tweets by AntiVaxxerMom will be retrieved. If that account has replied to any other messages in the extant corpus that are flagged as vaccine relevant, those tweets will be included in the corpus. This would include replies to the original conversation, and to other tweets in the corpus. In this way, continued conversation between users is captured, but other conversations that are otherwise unfindable involving those users are discovered as well. In addition, the remaining tweets by these users that were not involved in the conversation will be inspected to see if they are independently vaccine-relevant.

Note that this is an imperfect way to capture conversations. If user "AntiVaxxerTroll" replies to a message in any of these threads, but no other user involved in the conversation replies back to them, this would not be found. Also, because twitter limits timelines to 3,200 tweets, it is likely that especially prolific twitter users will not have the full time-period available.

\subsection{Code for Extraction and Analysis of Data}
The programming for retrieval and analysis of the data is being written, and is hosted in a public repository, github, with the exclusion of the cryptographic tokens and other private information needed to access the data, as well as the data itself, which twitter's TOS requires be kept private. (The message IDs are not required to be kept private, and will hopefully be shared once the corpus is constructed.)

\section{Metrics, Analysis Plans, and Tentative Hypotheses}

The primary descriptive metrics used to assess account behavior are being developed.

We expect that tweets in the corpus of conversations will be polarized due to the by-default combative nature of twitter interactions and groups\cite{Sunstein2002,Yardi2010}, and the increasing polarization of this debate specifically\cite{Witteman2012} shown in analyses of other platforms like Facebook\cite{Schmidt2018}\footnote{Getting past this polarization is a key goal, per Brunson \cite{Brunson2017}, but this is addressed by considering how risk communication is done, rather than by ignoring the clear current polarization.}. Given this, we expect the at pro- and anti- vaccine participants in the discussion will differ in their use of the key terms mentioned above. Specifically, the hypothesis is that pro-vaccine accounts will mention the identified health education terms first in conversations, while anti-vaccine accounts will mention the identified misinformation terms first in discussion. (This hypothesis may be checked based on whether it applies in a manual classification of a subset of accounts, or based on the manual classification of an external dataset.)

It may be difficult to automatically categorize the participants into pro-and anti-vaccine with high accuracy based purely on terms used, but prior work on this topic will be leveraged, or manual tagging of a selected subset can allow machine learning classification models for to be used. From this subset, features will be extracted. Features are likely to include; links to known (pro- or anti- vaccine) domains, terms from the earlier term list... (Need to think more about this. Perhaps I should chat with Doug Yeung?)

If available, the classification will be validated using the already classified sample from Broniatowski et al, 2019.

Multinomial Naive Bayes or logistic regression may be used based on the features list on a per-user basis to classify them into pro- or anti-vaccine. If these achieve greater than 90\% accuracy on a validation set, the results of this will be used. Otherwise, manual classification will be used for (a subset of) the corpus (if very large,) of conversations to provide (indicative) results.

\bibliographystyle{alpha}
\bibliography{Postdoc}

\end{document}
